Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-08-08 review/response.Rmd

3) Although the target of this article focuses on the dataset with two-dimensional structure (d=2), it is worth to discuss the possibility (or give some suggestions) of applying the proposed procedure to the situation where d > 2.

**We thank the referee for the suggestion. Many of the available index functions also work for d>2, for example the ones in the tourr package work for arbitrary dimension. The work in this paper focuses on studying a selection of indexes designed for 2D. However, the same diagnostics could be applied generally to any dimension projection index. We have commented on this in the discussion.**

4) The procedure for assessing the PPIs' properties was listed in Section 3.1.2. In addition, it would be helpful to list the procedure step-by-step (with the detailed R functions/packages) to guide the users how to apply the tour methodology with PPIs to explore the potential patterns of the real-world datasets.

**We have made a shiny app available in the package galahr, that will make it easier for new users to use a guided tour. The package documentation includes a step-by-step example. We have also added a step-by-step guide for the two applications in Section 4.3.**

@@

Thirdly, the simulations refer to sequences of projections (the horizontal axis in plots in Figures 2, 3, and 4. As I understand these sequences are a consequence of optimising the corresponding projection pursuit index. How can you then actually control the two different scenarios referred to in section 3.2 and 3.3? Or, do you extract a specified sequence of projections? Moreover, how do you make sure that you have the same sequence length for each trial? Please clarify!

**Captions and the text in Sections 3.2-3 have been changed to make it clearer that Figures 2 and 3 present the trace for a specified sequence of projections (since in the simulation study the nuisance and structured views are known), and Figure 4 is different as in this case the path is indeed the result of optimising an index.**

The optimisation methods are further discussed in section 3.4 and depend on the structure present in the data. What is the value of these investigations for more realistic settings in which the structure is not known.

**It is important to understand the performance of the optimisation before applying the method on new examples and we use the known examples for this. This is now explained in the first sentence in Section 3.4.**

@@

Finally, the relationship between the simulation study and the examples using the astrophysics data remains unclear. How are the physics applications linked to the simulation study and vice versa? Currently, the two aspects seem to me rather independent and could actually be presented in separate manuscripts. Moreover, since some rather important aspects of the simulation study have been put into an appendix. For the COST audience certainly the details of the simulation study are more relevant than the astrophysics application.

**The connection between two main sections is now explained in the introduction. We've added step wise instructions for applying projection pursuit to these two examples that can be used by researchers wishing to apply the techniques to their data.**

@@

Minor remarks
When introducing the various PPIs some terms and notations are used which are not properly introduced, e.g. alpha hull, convex. Check which ones actually need to be properly defined as they are needed for further steps and for which ones you could just refer to the literature avoiding unnecessary details.

**We have added more explanation, and appropriate references.**

### 2019-08-08 scagnostics.Rmd
final smoothing

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to too many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behaviour of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimization routines, and visualization.

\textcolor{red}{This paper first investigates the behaviour of these indexes to understand the potential and challenges in using them in projection pursuit settings, before applying them to examples of the physics problems we are interested in addressing.}
Section \ref{sec:construct} discusses index construction, and how they fit into the guided tour. Section \ref{sec:investigate} investigates the behaviour of the indexes, explored primarily using tour methods (@As85, @BCAH05, @CBCH94). The new guided tour with these indexes is applied to two examples from gravitational wave astronomy (Section \ref{sec:phys}).

@@

## Application guide

Applying projection pursuit to new datasets can be done using the guided tour available in the `tourr` package. The necessary steps are:

1. Scaling and standardization of the data
2. Index function selection and if necessary formatting according to the guided tour requirements, i.e. a function taking matrix input and returning the index value
3. Calling the guided tour with the data and index function:
    + for exploration this can be done via the `tourr::animate` function
    + for recording the results `tourr::save_history` should be used
4. Exploring how the results depend on choices made (index function, starting planes, optimisation method, reducing dimensionality)

These four steps are followed in the applications presented here and are documented in the comments of the source code [@paperSRC]. For simple usage examples see documentation of the `tourr::guided_tour` function.

@@

This article was created with R Markdown [@rmarkdown], the code for the paper is available at [@paperSRC].
Methods for testing new index functions as presented in this work are implemented in the R package spinebil [@spinebil].
The R package galahr [@galahr] provides a graphical interface to the tourr package allowing for interactive exploration using the guided tour.
Additional information is available in the appendix, covering details of how the holes index was rescaled, estimations of the squint angle, a comparison of the computational performence of the index functions, testing the effect of index parameters on the results and suggestions how the index functions may be refined. Finally additional figures are also presented in the appendix.

>>>>

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to too many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behaviour of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva].

Projection pursuit can be understood and investigated more substantially when combined with a tour. A tour (@As85, @BCAH05) displays a smooth sequence of low dimensional projections from high dimensions to explore multivariate data for structure such as clusters, outliers, and nonlinear relationships. @CBCH94 provided an approach combining the tour algorithm with projection pursuit, to interactively both search for interesting projections, and examine the behavior of the indexes. The projection pursuit guided tour is available in the R package, tourr [@tourr], and provides optimization routines, and visualization.

\textcolor{red}{This paper is structured as follows.} Section \ref{sec:construct} discusses index construction, and how they can be used in the guided tour. Section \ref{sec:investigate} investigates the behaviour of the indexes, explored primarily using tour methods. The new guided tour with these indexes is applied to two examples from gravitational wave astronomy (Section \ref{sec:phys}). \textcolor{red}{The latter two parts are connected in that the application of the new indexes to these problems is the main motivation for the paper, and the simulation study, in the first part, was conducted to better understand the behavior of the indexes in general. The techniques in Section \ref{sec:investigate} define procedures that will be generally useful for researchers developing new projection pursuit indexes to assess their behavior. The paper finishes with a discussion about the limitations of this work, and the potential future directions.}

@@

## Instructions for applying to new data

Applying these procedures to new datasets can be done using the guided tour available in the `tourr` package. The typical steps required are:

1. Scale, standardize or sphere (principal components) the data. 
2. Index function selection matching the type of structure that is interesting to detect. Any new function can be used as long as it takes a matrix as input and returns a single index value.
3. Call the guided tour with the data and index function:
    + for exploration this can be done via the `tourr::animate` function
    + for recording the results `tourr::save_history` should be used
4. Explore how the results depend on choices made (index function, starting planes, optimisation method, prior dimension reduction).

These are steps followed in the above two applications and are documented in the comments of the source code [@paperSRC]. For simple usage examples, see documentation of the `tourr::guided_tour` function.

@@

- This article was created with R Markdown [@rmarkdown], the code for the paper is available at [@paperSRC].
- Methods for testing new index functions as presented in this work are implemented in the R package spinebil [@spinebil].
- The R package galahr [@galahr] provides a graphical interface to the tourr package allowing for interactive exploration using the guided tour.
- Additional explanations are available in the Appendix, covering details of 
    - how the holes index was rescaled, 
    - estimations of the squint angle, 
    - a comparison of the computational performence of the index functions, 
    - testing the effect of index parameters on the results and 
    - suggestions how the index functions may be refined.
